Hospitalizations for psychiatric conditions among adolescents during the COVID-19 pandemic

### 2021-08-18 psy_peds.Rmd
adding percentage patients trends & icd_categories comparison

### Data management for the patientClinicalCourse table
```{r}
clinical_raw <- clinical_raw %>% 
  dplyr::mutate( date = as.Date( calendar_date, format = dateFormat ),
                 weeks = as.Date(cut( date, breaks= "week")),
                 year = format( date, "%Y"), 
                 period = ifelse( date < pandemic_start_date,
                                  "before_pandemic","during_pandemic")) %>% 
  dplyr::filter(cohort  %in% c("AllAdm", "AllNotAdm"))
```

@@

all_codes <- obs_raw %>% 
  full_join( clinical_raw,  by = c("patient_num", "cohort", "days_since_admission")) %>%
  full_join( demo_raw, by = c("patient_num", "cohort")) %>%
  dplyr::filter( concept_type == "DIAG-ICD10", 
                 in_hospital == 1,
                 age < 18 & age > 10,
                 date >= history_start_date & date <= history_end_date) %>%
  dplyr::select( patient_num, cohort, sex, age, concept_type, concept_code, calendar_date,
                 in_hospital, n_hospitalisation, len_hospitalisation, date, weeks, year,
                 period, disorder_group, description )

input_data <- all_codes %>% 
  filter(concept_code %in% icdCodes$ICD10_Code)

@@

patient_count_barplot_period$orderDisorder <- factor(patient_count_barplot_period$disorder_group,levels = as.factor( order_disorders$disorder_group ))

@@

## Diagnoses count per week
Here we calculate the number of patients hospitalized per week (exactly, the number of patients being assigned a diagnosis on that week, hence usually corresponding to the end of hospitalization, ie not taking into account the length of the encounter). 
The barplot shows the weeks on the x-axis, and the patient counts on the y-axis. The horizontal lines is the (quadratic) temporal trend of these patient counts.
```{r}
patient_count_per_week <- input_data[, c("patient_num", "cohort", "weeks")] %>%
  unique() %>%
  group_by(weeks) %>%
  tally() %>%
  mutate(n = ifelse( n < obfuscation | isFALSE(obfuscation),
                     n,
                     0.5),
         siteid = site )

#estimate the number of patients per week
patient_count_per_week %>%
  dplyr::mutate(weeks = as.Date(weeks)) %>% 
  dplyr::mutate(period = ifelse( weeks < pandemic_start_date,
                                  "before_pandemic",
                                 "during_pandemic")) %>% 
  ggplot(aes(x = weeks, y = n, fill = period)) +
  geom_bar(stat="identity")+
  scale_fill_manual(values = cbPalette) +
  labs(y = "Counts",
       x = "Calendar Weeks",
       title = "Weekly patient counts with mental health related ICD codes") + 
  geom_smooth(method = "lm", formula = y ~ splines::ns(x, 2), se = TRUE, aes(colour=period), colour= "black", size = 0.5)

@@

```{r}
patient_count_per_week <- all_codes %>% 
  mutate(icd_code_category = ifelse(concept_code %in% icdCodes$ICD10_Code, 
                               "psy", 
                               "somatic")
  ) %>% 
  select(patient_num, cohort, weeks, icd_code_category, period) %>% 
  unique() %>%
  group_by(weeks, icd_code_category, period) %>%
  tally() %>%
  mutate(n = ifelse( n < obfuscation | isFALSE(obfuscation),
                     n,
                     0.5),
         siteid = site )

@@

patient_count_per_week_byCategory <- left_join( patient_count_per_week_byCategory, icdCategory) %>%
  select(patient_num, cohort, weeks, icd_description_category, period) %>% 
  unique() %>%
  group_by(weeks, icd_description_category, period) %>%
  tally() %>%
  mutate(n = ifelse( n < obfuscation | isFALSE(obfuscation),
                     n,
                     0.5),
         siteid = site )

patient_count_per_week_byCategory %>%
  filter( icd_description_category %in% c("Neoplams","Diseases of the respiratory system", "Diseases of the nervous system", "Diseases of the digestive system", "Diseases of the blood", "Congenital malformations, deformations and chromosomal abnormalities" )) %>%
  ggplot(aes(x = weeks, y = n, fill = icd_description_category, color = icd_description_category)) +
  geom_point() +
  geom_line() +
  geom_vline(xintercept = as.Date(pandemic_start_date),
             linetype = "dashed") +
  scale_fill_manual(values = cbPalette) +
  scale_color_manual(values = cbPalette) +
  labs(y = "Counts",
       x = "Calendar Weeks",
       title = "Weekly patient counts") + 
  geom_smooth(method = "lm", formula = y ~ splines::ns(x, 2), se = TRUE, aes(colour=period), colour= "black", size = 0.5)

>>>>

### Data management for the patientClinicalCourse table
```{r}
clinical_raw <- clinical_raw %>% 
  dplyr::mutate( date = as.Date( calendar_date, format = dateFormat ),
                 weeks = as.Date(cut( date, breaks = "week")),
                 month = as.Date(cut( date, breaks = "month")),
                 year = format( date, "%Y"), 
                 period = ifelse( date < pandemic_start_date,
                                  "before_pandemic","during_pandemic")) %>% 
  dplyr::filter(cohort  %in% c("AllAdm", "AllNotAdm"))
```

@@

all_codes <- obs_raw %>% 
  full_join( clinical_raw,  by = c("patient_num", "cohort", "days_since_admission")) %>%
  full_join( demo_raw, by = c("patient_num", "cohort")) %>% 
  mutate(icd_code_category = ifelse(concept_code %in% icdCodes$ICD10_Code, 
                               "psy", 
                               "somatic")) %>% 
  dplyr::filter( concept_type == "DIAG-ICD10", 
                 in_hospital == 1,
                 age < 18 & age > 10,
                 date >= history_start_date & date <= history_end_date) %>%
  dplyr::select( patient_num, cohort, sex, age, concept_type, concept_code, calendar_date,
                 in_hospital, n_hospitalisation, len_hospitalisation, date, weeks, month, year,
                 period, disorder_group, description, icd_code_category )

input_data <- all_codes %>% 
  filter(concept_code %in% icdCodes$ICD10_Code) %>%

@@

patient_count_barplot_period$orderDisorder <- factor(patient_count_barplot_period$disorder_group,
                                                     levels = as.factor( order_disorders$disorder_group ))

@@

## Diagnoses count per week
Here we calculate the number of patients hospitalized per week (exactly, the number of patients being assigned a diagnosis on that week, hence usually corresponding to the end of hospitalization, ie not taking into account the length of the encounter). 
The barplot shows the weeks on the x-axis, and the patient counts on the y-axis. The horizontal lines is the (quadratic) temporal trend of these patient counts.

```{r}
all_codes[, c("patient_num", "cohort", "month")] 
```

```{r}
patient_count_per_week <- all_codes[, c("patient_num", "cohort", "weeks", "icd_code_category")] %>%
  unique() %>%
  group_by(weeks, icd_code_category) %>%
  tally() %>%
  mutate(n = ifelse( n < obfuscation | isFALSE(obfuscation),
                     n,
                     0.5),
         siteid = site )

patient_count_per_month <- all_codes[, c("patient_num", "cohort", "month", "icd_code_category")] %>%
  unique() %>%
  group_by(month, icd_code_category) %>%
  tally() %>%
  mutate(n = ifelse( n < obfuscation | isFALSE(obfuscation),
                     n,
                     0.5),
         siteid = site )

```{r}
#estimate the number of patients per week
patient_count_per_week %>%
  tidyr::pivot_wider(names_from = icd_code_category, 
                     values_from = n) %>% 
  dplyr::mutate(weeks = as.Date(weeks)) %>% 
  dplyr::mutate(period = ifelse( weeks < pandemic_start_date,
                                  "before_pandemic",
                                 "during_pandemic"), 
                percentage = psy * 100 / (somatic + psy)) %>% 
  ggplot(aes(x = weeks, y = percentage, fill = period)) +
  # geom_bar(stat="identity")+
  scale_fill_manual(values = cbPalette) +
  labs(y = "Counts",
       x = "Calendar Weeks",
       title = "Percentage of patients with mental health related ICD codes (weekly)") + 
  geom_smooth(method = "lm", formula = y ~ splines::ns(x, 2), se = TRUE, aes(colour=period), colour= "black", size = 0.5)

@@

```{r}
#estimate the number of patients per week
patient_count_per_week %>%
  dplyr::mutate(weeks = as.Date(weeks)) %>% 
  dplyr::mutate(period = ifelse( weeks < pandemic_start_date,
                                  "before_pandemic",
                                 "during_pandemic")) %>% 
  ggplot(aes(x = weeks, y = n, fill = icd_code_category)) +
  # geom_bar(stat="identity")+
  scale_fill_manual(values = cbPalette) +
  labs(y = "Counts",
       x = "Calendar Weeks",
       title = "Weekly patient counts with mental health related ICD codes") + 
  geom_smooth(method = "lm", formula = y ~ splines::ns(x, 2), se = TRUE, aes(colour=icd_code_category), colour= "black", size = 0.5)

```{r}
patient_count_per_week <- all_codes %>% 
  select(patient_num, cohort, weeks, icd_code_category, period) %>% 
  unique() %>%
  group_by(weeks, icd_code_category, period) %>%
  tally() %>%
  mutate(n = ifelse( n < obfuscation | isFALSE(obfuscation),
                     n,
                     0.5),
         siteid = site )

@@

patient_count_per_week_byCategory <- left_join( patient_count_per_week_byCategory, icdCategory) %>%
  mutate( icd_description_category = ifelse( concept_code %in% icdCodes$ICD10_Code, "psy", icd_description_category)) %>%
  select(patient_num, cohort, weeks, icd_description_category, period) %>% 
  unique() %>%
  group_by(weeks, icd_description_category, period) %>%
  tally() %>%
  mutate(n = ifelse( n < obfuscation | isFALSE(obfuscation),
                     n,
                     0.5),
         siteid = site )

patient_count_per_week_byCategory %>%
  filter( icd_description_category %in% c("psy","Neoplams","Diseases of the respiratory system", "Diseases of the nervous system", "Diseases of the digestive system", "Diseases of the blood", "Congenital malformations, deformations and chromosomal abnormalities" )) %>%
  ggplot(aes(x = weeks, y = n, fill = icd_description_category, color = icd_description_category)) +
  #geom_point(alpha=0.2) +
  geom_line() +
  geom_vline(xintercept = as.Date(pandemic_start_date),
             linetype = "dashed") +
  scale_fill_manual(values = cbPalette) +
  scale_color_manual(values = cbPalette) +
  labs(y = "Counts",
       x = "Calendar Weeks",
       title = "Weekly patient counts") +   
  theme(axis.text.x = element_text(angle=45, vjust = 1, hjust=1), legend.position = "bottom", legend.text = element_text(size=7), legend.title = element_text(size=7) ) +
  geom_smooth(method = "lm", formula = y ~ splines::ns(x, 2), se = TRUE, aes(colour=period), colour= "black", size = 0.5)
